---
title: "TEAM_PROJECT"
author: "Bryan Gutierrez"
date: "2025-02-19"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





```{r}
# read the csv
data <- read.csv("/Users/bryangutierrez/Downloads/mobility-all.csv", stringsAsFactors = FALSE) 


```



# Research Question 1 



```{r}

# I asked chatgbt how to see what variables are good to see economic mobility it recomended doing a coralation analysis 

# Load necessary library
library(tidyverse)

# Select only numeric variables
numeric_data <- select(data, where(is.numeric))

# Compute correlation with Mobility
cor_results <- cor(numeric_data, use = "complete.obs")["Mobility",]

# Sort correlations in descending order
cor_results <- sort(cor_results, decreasing = TRUE)

# see top correlated variables
cor_results
```






```{r}
library(tidyverse)
library(reshape2)

# I did this chart ecause i feel like it offers another view and we can see multicolinarity paterns were red is grouoped up

# Select numeric columns
numeric_data <- select(data, where(is.numeric))

# Calculate correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Melt the correlation matrix
melted_cor <- melt(cor_matrix)

# Plot the heatmap with improved formatting
ggplot(data = melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6)) +
  coord_fixed() +
  labs(title = "Correlation Heatmap for Economic Mobility",
       x = NULL, y = NULL)

```
Important Positive Correlations- Middle_class (0.67) and Commute (0.63) suggest that higher middle-class representation and longer commutes may be linked to higher mobility.

Important Negative Correlations- Single_mothers (-0.67) and Gini_99 (-0.64) suggest that inequality and single-parent households might limit mobility.




```{r}

#Regressional Model Multivariate 
numeric_columns <- sapply(data, is.numeric)
data_numeric <- data[,numeric_columns]

full_model <- lm(Mobility ~ ., 
                 data = data_numeric)

summary(full_model)



```



The variables that stood out to use as significant predictors for economic mobility.

Positively- Commute(p= 0.0049), Middle_class(p = 0.0393), Progressivity(p = 3.78e-07), Religious(p=2.81e-07)

Negatively-Black (p = .0003), Seg_racial (p = .0051), HS_dropout(0.0237), Manufacturing(p=4.52e-11), Violent_crime(p = .0372)




```{r}
# important variables that are imoortant with coralations and the regression 

refined_model1 <- lm(Mobility ~ Middle_class + Commute + Single_mothers + 
                    Seg_racial + Local_tax_rate + Religious, data = data)

summary(refined_model1)

```
```{r}
# Plotting residual diagnostics to check the models details 
par(mfrow=c(2,2))  # 2x2 grid
plot(refined_model1)  

```


```{r include=FALSE}
# Select only numeric columns
numeric_columns <- sapply(data, is.numeric)
data_numeric <- data[, numeric_columns]

# Remove 'Mobility' from predictor list
predictors <- setdiff(names(data_numeric), "Mobility")

# Initialize an empty list to store results
results <- list()

# Loop through each predictor
for (var in predictors) {
  # Run bivariate regression
  formula <- as.formula(paste("Mobility ~", var))
  model <- lm(formula, data = data_numeric)
  
  # Store summary in results list
  results[[var]] <- summary(model)
}

# Display results
results

```





```{r}


# Create an empty data frame to store results
bivariate_results <- data.frame(Variable = character(),
                                Estimate = numeric(),
                                P_Value = numeric(),
                                R_Squared = numeric(),
                                stringsAsFactors = FALSE)

# Loop through each predictor for bivariate regression
for (var in predictors) {
  formula <- as.formula(paste("Mobility ~", var))
  model <- lm(formula, data = data_numeric)
  model_summary <- summary(model)
  
  # Append results to the data frame
  bivariate_results <- rbind(bivariate_results, data.frame(
    Variable = var,
    Estimate = coef(model_summary)[2, "Estimate"],
    P_Value = coef(model_summary)[2, "Pr(>|t|)"],
    R_Squared = model_summary$r.squared
  ))
}

# Display results as a formatted table
library(knitr)
kable(bivariate_results, caption = "Bivariate Regression Results")



```




# Research Question 2

## Variable selection process

To investigate the extent to which measures of better education predict higher levels of economic mobility, I chose variables revolving around performance indicators and resource factors in regards to education.

**Educational Performance Indicators:** Test_scores, HS_dropout, Graduation **School Resource Factors:** School_spending, Student_teacher_ratio


```{r}
model <- lm(Mobility ~ Test_scores + HS_dropout +
              Graduation + School_spending +
              Student_teacher_ratio, data = data)
```





```{r}
# Cleaning up model by getting rid of outliers
cooksd <- cooks.distance(model)
influential_points <- which(cooksd > (4/length(cooksd)))
data_clean <- data[-influential_points, ]
model_clean <- lm(Mobility ~ Test_scores + HS_dropout + Graduation + 
                  School_spending + Student_teacher_ratio, data = data_clean)
summary(model_clean)
```

## Variable Interpretation

**Test_scores (0.0025, p \< 0.001):** Higher test scores are positively associated with greater economic mobility. **HS_dropout (-0.5690, p \< 0.001):** Higher high school dropout rates significantly decrease economic mobility. **Graduation (-0.0347, p = 0.0084):** Higher graduation rates negatively correlate with mobility, suggesting that other factors may be influencing this relationship. **School_spending (0.0025, p = 0.1506):** While increased school spending appears to have a positive effect, it is not statistically significant at the 5% level. **Student_teacher_ratio (-0.0010, p = 0.2969):** The student-to-teacher ratio does not significantly impact economic mobility,

## Model Diagnostics and Selection

```{r}
# Model without Student_teacher_ratio
reduced_model <- lm(Mobility ~ Test_scores + HS_dropout + Graduation + School_spending, data = data_clean)

```

**Test_scores (0.0025, p \< 0.001):** Higher test scores are positively associated with greater economic mobility. **HS_dropout (-0.5690, p \< 0.001):** Higher high school dropout rates significantly decrease economic mobility. **Graduation (-0.0347, p = 0.0084):** Higher graduation rates negatively correlate with mobility, suggesting that other factors may be influencing this relationship. **School_spending (0.0025, p = 0.1506):** While increased school spending appears to have a positive effect, it is not statistically significant at the 5% level. **Student_teacher_ratio (-0.0010, p = 0.2969):** The student-to-teacher ratio does not significantly impact economic mobility,

## Model Diagnostics and Selection

```{r}
# Model without Student_teacher_ratio
reduced_model <- lm(Mobility ~ Test_scores + HS_dropout + Graduation + School_spending, data = data_clean)

```


```{r}

par(mfrow = c(1, 3))

# Residuals vs. Fitted Plot
plot(reduced_model$fitted.values, resid(reduced_model), 
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red", lwd = 2)

# Q-Q Plot
qqnorm(resid(reduced_model))
qqline(resid(reduced_model), col = "red", lwd = 2)

# Cook's Distance
plot(cooksd, type = "h", main = "Cook's Distance", ylab = "Cook's Distance")
abline(h = 4/(nrow(data)), col = "red")

```











# Research Question 3




## Variable Selection Process

To investigate the extent to which measures of integration across social groups predict economic mobility, I selected a set of variables that capture different dimensions of social integration and demographic characteristics. I grouped them info two groups:

- **Segregation Measures** : Seg_racial, Seg_poverty, and Seg_affluence
- **Social and Demographic Factors** : Married, Divorced, Foreign_born, and Religious


```{r}
 # Create Variables for Social Groups
 social_groups <- c("Seg_racial", "Seg_poverty", "Seg_affluence", "Mobility", "Married", "Divorced", "Religious", "Foreign_born")
 data <- data[social_groups] %>% na.omit()
```







```{r, echo=FALSE}
model1 <- lm(Mobility ~ ., data = data)
cooksd <- cooks.distance(model1)
outliers <- which(cooksd > (4/length(cooksd)))
data_clean <- data[-outliers, ]
final_model <- lm(Mobility ~ ., data = data_clean)
summary(final_model)
```

## Model Diagnostics
```{r, echo=FALSE}
par(mfrow=c(1,3))
 plot(final_model$fitted.values, residuals(final_model),
 main="ResidualsvsFitted",xlab="FittedValues",ylab="Residuals")
 abline(h=0,col="red",lwd=2)
 qqnorm(residuals(final_model))
 qqline(residuals(final_model),col="red")
 plot(cooksd,main="Cook'sDistance",ylab="Cook'sDistance")
 abline(h=4/length(cooksd),col="red")
```



### **1. Residuals vs. Fitted Plot**: This plot displays the residuals against the fitted values from the regression model. This model is assumed valid because the plots are scattered randomly.  

### **2. Normal Q-Q Plot**: This plot checks whether the residuals follow a normal distribution. It compares the observed residuals to a theoretical normal distribution. Becuase the points follow the red line, the residuals are normally distributed

### **3. Cook's Distance Plot**: This plot identifies influential data points in the regression model. Cook’s distance measures how much a single observation affects the model. In this model, I removed all outliers seen on the plot. 



# Research Question 4
```{r}
# read the csv
data <- read.csv("/Users/bryangutierrez/Downloads/mobility-all.csv", stringsAsFactors = FALSE) 


```
# Variable Selection Process

To investigate variables which can be directly affected by government policy predict economic mobility, we looked into the following:

- **Tax Policy & Government Spending** (Local_tax_rate), (Progressivity)
- **Education Policy** (HS_dropout)
- **Public Safety & Social Programs** (Violent_crime), (Seg_racial)
- **Community & Infrastructure** (Single_mothers), (Commute), (Religious)





```{r}

economic_model <- lm(Mobility ~ Local_tax_rate + Progressivity + HS_dropout + Violent_crime +
         
         Seg_racial + Single_mothers + Commute + Religious, data = data)

summary(economic_model)

```


